Percolation Analysis – Building a Package

### 2020-04-21 README.Rmd
readme is now old vignette

>>>>

## Percolation Analysis via the R package “percopackage” – Workflow and Descriptions

Percopackage is an R-package, which takes scripts developed by Simon Maddison and Elsa Arcaute to create three functions, which perform

  a) the actual percolation analysis (function called “percolate”), 
  b) maps the clusters created by percolation (function “mapClusters”) and 
  c) creates three different plots of the analysis results (function “plotClustFreq”)

The percolate function does the actual computing of the percolation analysis and exports different tables on which the following functions operate. 
Two different data inputs are possible: A set of points in Euclidian space, for which a distance matrix will be calculated OR a distance matrix, which may for example have been calculated using least cost paths. If a distance matrix is provided, the point set is still needed for the mapping function.

### Directory Structure
The functions create subdirectories of the working directory the user is working in (using getwd() ), therefore we describe here the relative structure, which is being set up:

  * /analysis_results – output plots from the analysis, now principally the cluster size vs radius plots and the csv of the analyis results (described further down)

  * /maps – outputs from the mapping, showing clusters and nodes overlaid on given outline shape files, as well as output shape files with the point data (shape file called “map_name.shp”)

  * /working_data – working data generated by the programs and used to pass intermediate data between them.

### Percolate - function
The input needed for the percolate-function is as follows:

_percolate(data, distance_table = NULL, limit, radius_unit, upper_radius, lower_radius, step_value)_

**data** = A data.frame, with the format below:

PlcIndex,Easting,Northing

1,350350,233050

2,354700,260200

3,358700,238900

*PlcIndex* is an assigned index number to distinguish the points. This does not need to be sequential and can for example be drawn from a source database.

*Easting* and *Northing* are self-explanatory, and are to the metre. (Note that additional columns, if any, will be ignored).

**distance_table** is set to be NULL as a default, which means, from the data input (coordinates) a distance matrix is calculated using Euclidean distance. If not NULL, the data frame given here needs to adhere to this format:

ID1, ID2, d12

1,2,12.1

1,3,14.2

2,3,2.9

*ID1* and *ID2* are point IDs (i.e. the PlcIndex values) and

*d12* denotes the weighted distance between them, if a non-euclidian distance is whished to be used.

**limit** is an integer, the value above which distances will not be calculated between sites.

**radius_unit**  is either 1 for meter or 1000 for km for all input radii.

**upper_radius** is an integer, the upper value of the radius range to be used.

**lower_radius** is an integer, the lower value of the radius range to be used.

**step_value** is numeric or integer and the step value to be used between lower and upper radius.

**Exported** are in the folder *working_data*: tables (csv) with the input-data, the list of PlcIndex, null_entries and duplicate_entries to be used by following functions.

In the folder *analysis_results* the tables (csv) analysis_by_radius and member_cluster_by_radius are exported. They will be used by the following functions as well, but may be useful for other applications as well, therefore they are considered analysis results.

#### Note 
The function creates the Inter-site distance matrix – given the dataset of sites and XY coordinates, this computes a partial matrix of inter-site distances. The limit is set by the **limit** parameter in the function call. The inter-site distance is computed using Pythagoras’ theorem. The output file is located in */working_data* and is called: nodes_list_d.txt. See below for examples and explanation of the limit parameter.

It also creates a file listing the nodes that have been used in creating this list, which omits duplicates and null values etc. This is called PlcIndex.csv.

Cluster extraction is the second step: This program identifies all clusters for a specific radius. It steps through a range of radii, the range being set by input-parameters. The data is output as a text csv file called: member_cluster_by_radius.csv. This program is the heart of the analysis; the method of creating a graph where nodes are connected when the distance between them is the given radius or less, then extracting each cluster, remains as written by Elsa Arcaute.

Note that for large numbers of points, and large number of selected radius values, the computation time can be quite long. We recommend experimenting initially with only a small number of steps, e.g. with a small radius range, or with a large step value, in order to see which range of values is the most interesting.

For Hillforts in Britain, Simon Maddison ran the analysis with 1 km steps between 2 and 40km, but also ran an analysis for 3-6km with 0.1km steps to investigate clusters in high density areas of Scotland and Wales. To run with 0.1km steps over the full range would not only take excessive time but also produce large numbers of very similar plots.

Duplicate and null entries are also listed in files with these names, but should have no values if the data is clean. (Historically there was a lot of cleaning to do, which is why this runs as a separate program).

#### Example:

*percolate(data = database,, limit = 50, radius_unit = 1000, upper_radius = 40, lower_radius = 2, step_value = 1)*

A coordinates dataset is the database input. In the example above, percolation analysis will run for radius values between 40 and 2 km, with a step value of 1km (determined by the radius_unit parameter, set to 1000). When computing the inter-site distances, values greater than 50km will not be stored. This restricts the intermediate file size and improves performance. (There is no point in computing the distance between sites in Scotland and Cornwall for example, when all sites are within a single cluster above a radius of say 40km).

### mapClusters - function

This function will map the clusters created by percolation. It needs the following input:

*mapClusters(shape, map_name, source_file_name, dpi=300)*

**shape** is a SpatialPolygonDataFrame (sp), which will be used for the plotting as a background map, and determining the coordinate system.

**map_name** is a string, which will be used to title the maps.

**source_file_name** is a string, which will be used to label the maps and describes the source data used.

**dpi** is the output file resolution. The default is set to 300.

**Exported** maps are in the folder /maps: As many png map files as there are different radii values used in the percolation analysis. Also a shp-file with the point data.

#### Note
Mapping clusters – this function maps clusters for each radius on an outline map overlay. The same data can for example be mapped on a coastal outline, an outline with modern counties, or an outline with historic counties. The outline is defined by an ESRI shape file, which needs to be read, and passed to the function as an appropriate data frame. The outputs generated were originally a multi-page pdf file, but this proved unwieldy, and it was changed to generate png files, which are much easier to import into Microsoft Word and other applications, as well as to display. Point data is also exported as an ESRI shape file for import directly into ArcGIS etc.  Outputs are placed in the /maps sub-directory.

Clusters are ranked in order of size for each different threshold radius. The top 15 clusters are assigned a colour code with Red the largest, Blue the next and so on. A legend is plotted with the assigned cluster number against the colour. The plots also include the map name and the name of the source file used to generate the data, as well as the radius value.

The map projection is extracted from the given shape file and used for configuring the point plots and output shape files. This has not been extensively tested for different projection/ coordinate systems, and may need further development/modification for different input shape files. If necessary experiment by omitting the shape file plot by commenting out, and simply plot the points, in the first instance. Points and shp need to have the same coordinate system and projection.

#### Example

*mapClusters(data = database, shape = outline_GB, map_name = “Britain Iron Age Hillforts”, source_file_name = “Hillfort Sites”)*

The output-maps will be called e.g. “percolation_plots_Britain Iron Age Hillforts_rad_40.png” (number changing regarding to which percolation radius is depicted). On top of the map, the title will be: “Britain Iron Age Hillforts” and it will have a subtitle with a string like: “Source File: Hillfort Sites, percolation distance: 40 km”.

## plotClustFreq - function

Percolation cluster size and rank analyses – this program processes the cluster data generated by the cluster extraction program to give various different frequency and ranking plots.

*plotClutsFreq(source_file_name)*

**source_file_name** is a string, which will be used to label the plots and describes the source data set. If already input in mapClusters this doesn’t need to be done again.

Three png-files are generated:

  a) radius to maximum cluster size, 
  b) radius to mean cluster size and 
  c) radius to normalized max. cluster size

and they will include the information of the source file as a subtitle.

At the moment they are stored in the /analysis_results directory as “radius_to_max_cluster_size”, “radius_to_mean_cluster_size” and “radius_to_norm_max_cluster_size”.

#### Note

This program evolved significantly as various different statistical analyses were experimented with, under the guidance of Mark Lake. The final version reflects the latest work by Elsa Arcaute. The outputs are placed in the /analysis_results sub-directory.
The most useful plot is the cluster transition plot, showing the maximum cluster size (normalized) vs. radius, which shows the transitions.

### CONTACT DETAILS:
Simon Maddison
simon.maddison@btinternet.com
+44 797 089 2171

Sophie C. Schmidt
s.c.schmidt@uni-koeln.de

# Bibliography

Arcaute, E., Molinero, C., Hatna, E., Murcio, R., Vargas-Ruiz, C., Masucci, A P & Batty, M 2016 Cities and regions in Britain through hierarchical percolation. Royal Society Open Science, 3 (4): 150691.DOI: 10.1098/rsos.150691

Rozenfeld, H. D., Rybski, D., Andrade, J. S., Batty, M., Stanley, H. E. & Makse, H. A. 2008, Laws of population growth. Proceedings of the National Academy of Sciences of the United States of America, 105 (48): 18702.DOI: 10.1073/pnas.0807435105

Rozenfeld, H. D., Rybski, D., Gabaix, X. & Makse, H. A. 2011, The Area and Population of Cities: New Insights from a Different Perspective on Cities. American Economic Review 101 (5): 2205-25.
